Supplementary material

### 2023-09-02 manuscript/submission2/outbreak_response_models_review_v2.tex

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Modelling outbreak response impact in human vaccine-preventable diseases: A systematic review of differences in practices between collaboration types before COVID19}
}
\newline
\\
James M. Azam\textsuperscript{1, 2*},
Xiaoxi Pang\textsuperscript{3},
Elisha B. Are\textsuperscript{2, 4},
Juliet R.C. Pulliam\textsuperscript{2},
Matthew J. Ferrari\textsuperscript{5}
\\
\bigskip
\textbf{1} Department of Infectious Disease Epidemiology, London School of Hygiene and Tropical Medicine, London, United Kingdom
\\
\textbf{2} DSI-NRF Centre of Excellence in Epidemiological Modelling and Analysis (SACEMA), Stellenbosch University, Stellenbosch 7600, South Africa
\\
\textbf{3} Department of Mathematics, The University of Manchester, Manchester, United Kingdom
\\
\textbf{4} Department of Mathematics, Simon Fraser University, Burnaby, BC, Canada
\\
\textbf{5} Center for Infectious Disease Dynamics, Department of Biology, The Pennsylvania
State University, University Park, Pennsylvania, USA
\\
\bigskip

* james.azam@lshtm.ac.uk

\end{flushleft}
\section*{Abstract}
\subsection*{Background}
Outbreak response modelling often involves collaboration among academics, and experts from governmental and non-governmental organisations. We conducted a systematic review of modelling studies on human vaccine-preventable disease (VPD) outbreaks to identify patterns in modelling practices between two collaboration types. We complemented this with a mini comparison of foot-and-mouth disease (FMD), a veterinary disease that is controllable by vaccination.   
\subsection*{Methods}
We searched three databases for modelling studies that assessed the impact of an outbreak response. We extracted data on author affiliation type (academic institution, governmental, and non-governmental organizations), location studied, and whether at least one author was affiliated to the studied location. We also extracted the outcomes and interventions studied, and model characteristics. Included studies were grouped into two collaboration types: purely academic (papers with only academic affiliations), and mixed (all other combinations) to help investigate differences in modelling patterns between collaboration types in the human disease literature and overall differences with FMD collaboration practices. 
\subsection*{Results}
Human VPDs formed 227 of 252 included studies. Purely academic collaborations dominated the human disease studies (56\%). Notably, mixed collaborations increased in the last seven years (2013 - 2019). Most studies had an author affiliated to an institution in the country studied (75.2\%) but this was more likely among the mixed collaborations. Contrasted to the human VPDs, mixed collaborations dominated the FMD literature (56\%). Furthermore, FMD studies more often had an author with an affiliation to the country studied (92\%) and used complex model design, including stochasticity, and model parametrization and validation. 
\subsection*{Conclusion}
The increase in mixed collaboration studies over the past seven years could suggest an increase in the uptake of modelling for outbreak response decision-making. We encourage more mixed collaborations between academic and non-academic institutions and the involvement of locally affiliated authors to help ensure that the studies suit local contexts.

@@

\subsection*{Keywords}
Infectious disease dynamics; outbreak response modelling

\section*{Highlights}
\begin{enumerate}
	\item We explored differences in collaborative modelling practices in the human-vaccine preventable disease outbreak response literature.
	\item Collaborations were grouped into purely academic and mixed.
	\item The collaboration types differed in terms of their model choices and practices.
	\item Purely academic collaborations dominated the human vaccine-preventable diseases literature.
	\item Mixed collaborations were more likely to involve local experts and use complex methods.
\end{enumerate}

\linenumbers

@@

\section*{Introduction}
Successful outbreak response to infectious diseases is often the result of a highly collaborative process~\cite{Sigfrid2020}. Decision-making during outbreak response usually requires collaborations between academic and field experts, including governmental and Non-Governmental Organizations (NGOs)~\cite{Kretzschmar2020,Kerkhove2012,Okiror2021,Whitty2014,Sigfrid2020}. These collaborations ensure that important perspectives from both research and operations/implementation are accounted for in the decision-making process.

Outbreak response decision-making often requires adaptations to the affected location ~\cite{Whitty2014,Heesterbeek2015,Sigfrid2020}. The interventions and decisions are usually driven by data/evidence, information, and experiences from the past or in real-time, either locally or from similar phenomena elsewhere ~\cite{Abramowitz2015,Kerkhove2012}. Furthermore, involving local experts in the design of interventions has been found to boost the success of outbreak response efforts ~\cite{Sigfrid2020}. Consequently, outbreak response should ideally involve at least one local expert to provide more context~\cite{Sigfrid2020}.

Mathematical modelling can support outbreak response decision-making~\cite{Lofgren2015}. Modelling is a proven tool for revealing insights about the extent of disease spread, and impact of interventions, and providing recommendations for decision-making during outbreaks~\cite{Kretzschmar2020,Whitty2014,Lofgren2015}. Insights from mathematical modelling, though often useful, only form part of the larger context (socio-economic, political, and so forth) to be considered during an outbreak, making it difficult to determine the extent to which it contributes to outbreak response decision-making ~\cite{Kretzschmar2020,Kerkhove2012}. One way to help ensure that modelling contributes to decision-making is the conduct of interdisciplinary research  --- that is, research conducted by authors with diverse scientific backgrounds --- between model developers and decision-makers.

The link between interdisciplinarity in scientific research and research impact, for example, number of citations has been well-researched~\cite{Abramo2017,Wagner2011}. However, few studies have investigated the impact of interdisciplinary collaborations on the conduct of scientific research. One such study investigated the impact of interdisciplinarity on the scientific validity of the methods used in a selection of papers that applied machine learning on topics in biology or medicine~\cite{Littmann2020}. The study found that the methods used in the reviewed studies differed according to the nature of the scientific backgrounds of the authors who conducted the research.

Our preliminary search of the literature did not reveal any systematic reviews investigating differences in research practices among collaborations in the outbreak response modelling literature. We, therefore, conducted this systematic review of outbreak response modelling studies of human vaccine-preventable diseases with the aim to explore the time and geographic patterns in the collaboration landscape, and to investigate differences in model choices and modelling practices. We categorised the collaboration landscape into two types: (1) purely academic collaborations: papers with all authors affiliated to academic institutions, and (2) mixed collaborations, that is, papers with authors affiliated to a mixture of academic, government and NGO affiliations. Using this classification, we investigated differences in the choice of model structure, dynamics, individual and spatial heterogeneity, and modelling methods such as model parametrization, validation, sensitivity analysis, the use of data and provision of data and code for reproducibility.

We sought to answer the following questions with respect to the two collaboration types in the human disease outbreak response modelling literature: (1) How are the studies distributed in time (years) and geographic locations, (2) Is there a difference in the practice of including local experts in modelling exercises? (3) Do the choice of model characteristics (structure and dynamics) and modelling methods/approaches (model parametrization and validation, outcomes of interest, sensitivity analyses, use of data, and code availability) differ between the collaboration types?

Even though this review is about human VPDs, we acknowledge and include the modelling work done in the veterinary literature and in particular, for foot-and-mouth disease (FMD). Specifically, the outbreak response modelling effort resulting from the 2001 (FMD) outbreak in the United Kingdom has been a foundational example of the application of mathematical modelling to decision-making. Subsequent work based on this outbreak has been instrumental in the development of FMD outbreak intervention strategies around the world. We, therefore, include some relevant FMD references for comparison. Essentially, in this review, we study the differences in terms of collaboration types in the human disease modelling literature, and compare our observations to some parallel results from the FMD modelling literature. This is useful for investigating any differences in approaches and practices between the human and livestock literature and opens up opportunities for future work.

The following were the objectives of the systematic review:
\begin{enumerate}
	\item To characterise the nature of collaboration types through geographic space and time,
	\item To investigate if there were differences between the collaboration types regarding the inclusion of local experts,
	\item To report differences between the collaboration types in terms of the choice of model structure, dynamics and heterogeneity, and
	\item To investigate and report differences in modelling methods/approaches (model parametrization, validation, sensitivity analyses, code availability, and so forth) between collaboration types.
	\item To report on miscellaneous interesting aspects captured by each study including the disease, interventions, and outcomes studied.
\end{enumerate}

@@

\section*{Materials and methods}
We followed the 2020 Preferred Reporting Items for Systematic Reviews and Meta-analyses statement (PRISMA 2020)  to conduct this systematic review \cite{Page2021}.

\subsection*{Eligibility criteria}
The following definitions were used in the determining study eligibility:

\begin{itemize}
	\item ``Outbreak'': a new and sudden rise in the number of cases of a disease in a population, which when left uncontrolled, could lead to large scale geographic spread, AND 
	\item ``Outbreak response'': an intervention directly triggered by the outbreak of an infectious disease, AND
	\item ``Mechanistic models'': mathematical models that use an equation or system of equations to capture the biological mechanisms driving the transmission dynamics of the infectious disease at the level of an individual or population \cite{Lessler2016a,Reiner2013}, AND
	\item ``Outbreak intervention assessment'': a mechanistic model-based evaluation of the impact of an outbreak response intervention.
\end{itemize}

We included studies that used a mechanistic model to investigate the impact of interventions triggered by the outbreak of any of the human vaccine-preventable diseases listed by the World Health Organization as of 2019 (Table 1 in \nameref{S1_File}). Aside from the WHO list, we also included Ebola because at the time of conducting this review, a vaccine had been approved for outbreak response purposes. We also included foot-and-mouth disease for comparison as a candidate veterinary disease that is manageable by vaccination and has been well modelled in the past.

Studies were considered unique even if they had a duplicated author list or a slight variation in the author list, probably representing the same modelling group, so far as the content of the paper was different.

@@

\subsection*{Data extraction}
To extract the data from the studies, we used the data extraction questionnaire implemented with the KoboToolbox web application (Section 3 in \nameref{S1_File}). The questionnaire was first piloted with $10$ papers among the three reviewers, who worked in duplicate. All conflicts in terms of extracted data were resolved through discussion and mutual agreement.

@@

Following the pilot phase, the full list of included studies was shared among three reviewers (JMA, XP, and EBA), who extracted the relevant data independently. The extracted data was cross-checked by one reviewer (JMA). When a discrepancy was found in the extracted data, the reviewer referred to the original paper and resolved the conflict.

From each paper, we extracted data on the type of publication (original research, preprint, and so forth), author affiliation type (academic/government/non-governmental), country/setting studied, whether at least one author affiliation was situated in the country studied, disease studied, study objectives (retrospective/prospective impact assessment, timing of the modelling practice with respect to the outbreak (retrospective/prospective/real-time), interventions studied, and whether vaccination was the most impactful intervention when compared as a single intervention (yes/no/inconclusive). A full list of the extracted data items are provided (Section 2 in \nameref{S1_File}).

From each model described in the methods section, we extracted data on the representation of individuals (compartmental vs individual-based), whether spatial structure was represented (yes/no), model structure (deterministic vs stochastic), model parametrization, and validation. We also extracted a predefined list of outcomes (final epidemic size, attack rate, cases averted, ad a host of others) measured with the models. If the outcome measured was not on the pre-defined list, we collected it separately as free text for further analysis. We collected data on whether the studies included sensitivity analysis or not. This was to understand how uncertainties in the model inputs were dealt with.  We collected data on whether the studies used observed data, that is, excluding simulated data, and whether the data was openly accessible. Lastly, we noted whether the code used for the analysis was openly available. In this regard, we classified code that was only available upon request as not available because of the barrier to access that this approach often introduces.

@@

To explore how the publications changed in time with respect to the collaboration types, we tallied the number of publications and relative proportion (mixed versus purely academic collaboration) of studies per year. We lumped together the publications from 1970 to 2005 due to the small numbers.

To study how the papers were distributed in terms of geographic locations, we split the studies into those that studied actual versus hypothetical locations and used the subset of studies about actual locations to rank the topmost studied locations (country and continent) by frequency.

We also investigated how connected the authors were to the locations studied. Here, we only considered studies on geographic scales up to actual countries and not larger. All studies about locations larger than a country were dropped from this part of the analysis. We tallied, in terms of collaboration types, the number of studies about actual locations (countries) with or without at least one author affiliation in the studied location. If a paper studied more than one location, we treated each location as a separate instance. This increased the denominators accordingly.

We summarized the interventions studied by grouping the studies into three categories, that is, studies that modelled non-vaccination interventions, those that modelled vaccination in combination with other interventions, and those that modelled vaccination as a single intervention for comparison with other interventions. We also summarised the top two most studied intervention for each disease by counting the number of times each disease was studied per paper.

For the disease studied, model characteristics (structure, dynamics, and spatial heterogeneity), and modelling practices (parametrization, validation, sensitivity analysis, data use and available, and code use and availability), we counted the number of studies per collaboration type and reported the results as percentages and fractions of the total. In counting the number of papers per disease studied and outcome measured, if a study had more than one outcome measured or disease, we treated each instance as unique and counted them separately. Hence, in such cases, the denominator of the reported fraction increased from the total studies we reported in our search results.

@@

To summarise the model outcomes, we stratified by the collaboration type and tallied the outcomes used within each group. We reported the top six most used outcomes by both collaboration types and compared the results between the two.

All analyses were perform in R version 4.3.0 ~\cite{R2023}.  The raw analysis of the results presented in the main text have been provided as a supplement (\nameref{S2_File}). We have also supplied as supplementary files a pdf of tables with the database of the included studies and extracted data (\nameref{S4_Table}). The R code and data for this analysis can be found in a public Github repository accessible at \url{https://github.com/jamesmbaazam/orv_modelling_review_epidemics}.

\section*{Results}
\subsection*{Study selection}
We retrieved a total of $12, 986$ bibliographic records from Scopus, PubMed, and Web of Science (Fig \ref{fig:fig1prismaflowchart}). Using Endnote and Rayyan, we identified and removed $7, 974$ duplicates, resulting in $5, 012$ unique records. We exported the unique records into Rayyan for screening in two stages.

Stage one involved screening titles and abstracts in duplicate by the three reviewers. After this stage, $4, 211$ studies were excluded. At stage two, we screened the full text of the remaining studies, and this led to $549$ exclusions.

After stage two, $252$ studies remained for the data extraction stage. Of the $252$ studies, $227$ were about human vaccine-preventable diseases and $25$ studies about foot-and-mouth disease (FMD). The FMD studies were not used in the main analysis but only for comparisons in the dedicated section for FMD.

\begin{figure}[h!]
	\includegraphics[scale=0.65]{figs/fig1_PRISMA_flowchart}
	\caption{\bf The PRISMA flow chart. Numbers described here include studies for both the human vaccine-preventable diseases and foot-and-mouth disease.}
	\label{fig:fig1prismaflowchart}
\end{figure}

\subsection*{Publications over time}
Overall, a few relevant papers were published between 1970-2005 $(4.0\%; 9/227)$, followed by a marked increase in publications until 2019 $(96.0\%; 218/227)$ (Fig \ref{publications_per_year}).

\begin{figure}[!h]	
 \includegraphics[scale=0.70]{figs/fig2_publications_per_year.png}
	\caption{\bf Number of publications per year (1970-2019).}
	\label{publications_per_year}
\end{figure}

@@

We first categorised the papers according to the unique combinations of author affiliation types (Table \ref{studies_per_author_affiliations_type}).

Overall, papers with all authors from academic institution affiliations were the most common (56.8\%; 129/227). Papers with author affiliations from academic and governmental organizations were the second most common (24.2\%; 55/227), and followed by those from academic and NGO affiliations (7.5\%; 17/227). The least common were those from governmental and NGO affiliations (0.9\%; 2/227).

As explained earlier in the methods section, for the analyses moving forward, we grouped the author affiliation combinations into two collaboration types: purely academic and mixed. There were more purely academic collaborations (56.8\%; 129/227) than mixed collaborations (43.2\%; 98/227).

\begin{table}[!h]
\centering
\caption{\bf Number of studies per unique combination of author affiliations (human diseases). Here, we also show the categorization of the author affiliation combinations into purely academic and mixed collaborations. Papers authored by purely academic collaborations have all authors being affiliated to academic institutions whereas those authored by mixed collaborations have authors being affiliated to a combination of academic institutions, government institutions, and NGOs or only one of the last two.}
\setlength\arrayrulewidth{1pt} 
\begin{tabular}{|l l c|}
\hline
\textbf{Collaboration type} & \textbf{Author affiliation combination} & \textbf{Number of publications} \\ \hline
Purely academic & academic only  & 129 \\ \hline
Mixed & academic + governmental   & 55 \\ \hline
Mixed & academic + NGO                 & 17  \\ \hline
Mixed & academic + governmental + NGO    & 14 \\ \hline
Mixed & governmental only                & 5 \\ \hline
Mixed & NGO only                       & 5  \\ \hline
Mixed & governmental + NGO               & 2 \\ \hline  \rowcolor{gray!20}
\textbf{Total} &  & \textbf{227} \\  \hline 
\end{tabular}
\label{studies_per_author_affiliations_type}
\end{table}

@@

To investigate the changes in collaboration types over time, we calculated the number and proportion of studies per collaboration type per year (Fig \ref{studies_per_collab_type}). In the past seven years (2013-2019), there was an absolute increase in the number of papers by both collaboration types. However, in the same period, there was no increase in the relative proportion of publications per year of mixed collaborations (\nameref{S1_Fig}).

\begin{figure}[!h]
\includegraphics[scale=0.65]{figs/fig3_total_collabs_per_year_plot.png}
	\caption{\bf Total studies by collaboration type. The period from 1970-2005 has been lumped up due to the lack of publications. Academic collaborations refer papers with all authors affiliated to academic institutions whereas mixed collaborations include papers authored by a mixture of authors who are affiliated with academic institutions, government institutions, and/or NGOs. Mixed collaborations are represented by the turquoise bars and purely academic collaborations by the brown bars.}
	\label{studies_per_collab_type}
\end{figure}

### 2023-10-18 analyses/scripts/S4_Table.Rmd
Save the tex versions as well

---
title: "Supplementary material"
output:
  pdf_document:
    fig_crop: no
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      dev = 'png', 
                      out.height = "\\textheight",  
                      out.width = "\\textwidth", 
                      message = FALSE, 
                      warning = FALSE
                      )

>>>>

---
title: "Supplementary material"
output:
  pdf_document:
    fig_crop: no
    keep_tex: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      dev = 'png', 
                      out.height = "\\textheight",  
                      out.width = "\\textwidth", 
                      message = FALSE, 
                      warning = FALSE
                      )
